%!TEX root = constraint-layout.tex
\section{Design of \projectname}

\projectname\ is a domain-specific language for the creation of highly
customized graph layouts based on properties of the nodes. Users identify
element sets from the graph and apply constraints to all elements in a
given set. These sets can be hierarchically composed to create a flexible,
nested layout. \projectname\ reduces specification effort and allows the
user to focus on the graph layout in terms of high-level properties of the
data rather on individual nodes. These layout specifications can also be
easily reused across graphs for which the nodes have the same properties,
which allows the user to apply a single layout design across multiple
graphs in the same domain. In this section, we discuss the design of
\projectname\ including the process for specifying sets and applying
constraints.

\jheer{I like that this section starts with a high-level description of the key concepts. However, I think it stops short, veering off to repeat earlier made claims about the benefits of the language. Instead, it might help to go further in laying out the core ideas. Namely: what is a ``set''? A crisp definition is needed. (Also not 100\% clear what an ``element'' is.) Then perhaps talk about constraints on/between sets and define the related terminology (collections?).}

\smallTreeExample

The main contribution of \projectname\ is the language for specifying
high-level constraints. \projectname\ specifications are composed of four
components: \emph{nodes}, \emph{links}, \emph{guides}, and
\emph{collections}. An example \projectname\ specification is shown in
Figure \ref{fig:small-tree-example}. Properties of the graph are included
as properties of the nodes and links. The user can then define collections
which are defined as a list of sets and a list of high-level constraints
that apply to the elements in each set of the collection. Users identify
\emph{sets} based on graph properties; sets can contain either graph nodes
or previously defined collections. Guides are an optional component and may
be used to define top-level layout boundaries (defined with an x and/or y
position), which can be referred to in the constraint specification.

\jheer{Why are nodes and links part of the spec? Those are just the data, but the spec should be applicable to any number of datasets. I understand why they are supported in your implementation (for convenience of testing and deployment among other things), but in a full language model the layout specification and data should be neatly separable.}

\todo{from Alan: what is the difference between a collection and a set? If
  none or not much, should we just use one term (maybe set)?}
\todo{Jane: the way I see it, the difference between a collection and a
set is in how it is being used at the time. A collection is used in the
constraint application stage to identify which sets should have constraints
applied to them. So collections apply constraints to sets, sets have
constraints applied to them. A collection can become a set (e.g. have
constraints applied to it in a downstream step). So there is a blurred line
of whether or not they are the same.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Specifying Sets in \projectname}

Each collection in \projectname\ is defined as a list of sets, where sets
are lists of elements. The elements in a set can be either nodes or
previously created sets, thus enabling a hierarchical composition of sets
for more complex layouts. In this language, we support several strategies
for specifying sets of nodes: \emph{partitioning}, \emph{expressions},
\orange{\emph{relations}}, and \emph{compositions}.

\jheer{Are collections and sets really *lists*, which by definition are ordered? I would think that, formally, sets are *sets*... I also find the terminology here confusing. An expression is just that: an expression. A set created by application of boolean predicate (predicate being a more specific technical term than expression!) should probably have a name other than ``expression''. Relation is also confusing, but maybe because I do a fair amount of database work where relation and relational algebra have a different meaning. Can relation be recast as an expression that makes reference to the graph structure, not just node attributes?}

\textbf{Partitioning.} Users can partition all graph nodes into sets by
identify a property (\todo{Partition by multiple properties?}) of the nodes
with which to group the nodes (see Figure \ref{fig:small-tree-example},
Line \texttt{17}). Nodes are then separated into disjoint sets based on the
partition property. The user may also specify lists of values that identify
which property values to \texttt{include} or \texttt{ignore} when
performing the partition. These properties allow the user to create
partitions where some nodes are excluded from the partition, and are
therefore not placed in any of the partition sets.

\textbf{Expressions.} For more flexibility in the definition of sets, users
can specify a concrete lists of sets to compute using arbitrary boolean
expressions. For each expression, each node is evaluated based on the user
defined expression to determine if it is included in the set. The user may
refer to properties of the node using dot syntax (e.g.,
\texttt{node.<property>}). Users may also optionally specify a
\texttt{name} for the set, which may be used by the composition set
specification. For this specification strategy, it is possible to create
node sets that are not disjoint, and may thus produce unsatisfiable
constraints. \todo{example figure}

\textbf{Relations.} For some sets, it may be useful to identify particular
relationships between nodes, which can be specified as a relation. For each
node, \todo{finish this section}

\contradictionExample

\textbf{Compositions.} Sets may also be formed through the hierarchical
composition of collections. For example, in Figure
\ref{fig:small-tree-example} we first create a collection of sets using
partition on Line \texttt{17}, named ``layer.'' On Line \texttt{22} we
create a new collection that contains the ``layer'' set. Line \texttt{28}
and \texttt{32} of Figure \ref{fig:contradiction-example} show the same
strategy, whereas Line \texttt{23} shows the shorthand composition for
producing the collection containing one set. For compositions, the user may
refer to any named entities previously defined in the specification (e.g.,
collections or named sets produced from \emph{expressions}).

\todo{``from'' property for set specification?}

When a named set is defined, the set element promotes the properties that
are equivalent for all elements in the set to a property of the set
element. For example, in Figure \ref{fig:small-tree-example}, when the user
defines a partition on the nodes based on the \texttt{depth} (Line
\texttt{17}), then each set element is given a \texttt{depth} property with
the value for that set. This property may now be referred to in other parts
of the \projectname\ specification (e.g., Line \texttt{23}).


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Applying Constraints in \projectname}

\label{sec:constraints}
Once a collection has been defined, users can apply layout constraints to
each of the sets in the collection. We identify four different types of
constraints that can be applied to set elements: \texttt{alignment},
\texttt{position}, \texttt{order}, \texttt{circle}, and \texttt{unit}. In
the following sections, we describe the behavior and properties of each of
these constraints.

% design section: rationale for why these are the right constructs (sufficiently expressive)
	% why are they user friendly?
	% hypotheses and reason hope/belief/evidence

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Alignment Constraints}
\constraint{\texttt{align} \emph{x} \texttt{axis}} Alignment constraints
specify the alignment of elements along an axis (Figure
\ref{fig:small-tree-example}, Line \texttt{18}). The constraint must
include the \texttt{axis} (e.g., \texttt{x} or \texttt{y}) on which the
elements should be aligned and may optionally include the
\texttt{orientation} for the alignment, one of: \texttt{center},
\texttt{left}, \texttt{right}, \texttt{top}, or \texttt{bottom}. When
applied to nodes, this constraint ensures that all the nodes are aligned
along the \texttt{axis}. When applied to sets, this constraint produces an
alignment based on the centroid or boundary of the set elements.

\emph{Rationale.} Alignment is a common characteristic seen when producing
graphs by hand \orange{citation} or utilizing particular layout strategies
\orange{citation} and is therefore crucial to support in \projectname.

\jheer{Why only x/y axes? Why not polar coordinates, for example? Or along arbitrary line segments defined as guides? I'm not saying you *should* support these, only that it's not clear why you focus where you do. Also I don't totally understand what alignment means here (nor what the default would be if unspecified). Readers may be confused about what you mean by applied to nodes vs. applied to sets, as earlier we talk about sets as the units that constraints are applied to. Also, centroid or boundary, which is used when and why?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Position Constraints}
\constraint{\texttt{position} \emph{right} \texttt{of}
  \emph{``top\_guide''}} Position constraints specify rules for the layout
of nodes relative to globally identified elements (\orange{figure}). The
constraint must identify a global element (e.g., a guide) to position the
set elements relative to and an orientation for the position (e.g.,
\texttt{left}, \texttt{right}, \texttt{above}, or \texttt{below}). The user
may optionally provide a \texttt{gap} for how far apart the elements should
be from the position element.

\emph{Rationale.} Position constraints allow the user to provide
overarching constraints relative to global elements such as guides. These
can be used to constrain the overall size of the graph or to section off
different areas. \todo{citations}

\jheer{Why only relative guides? Why can't relative position constraints be applied between sets?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Order Constraints}
\constraint{\texttt{order} \emph{y} \texttt{axis} \texttt{by}
  \emph{``depth''}} Order constraints enforce a sort order on the set
elements (Figure \ref{fig:small-tree-example}, Line \texttt{23}). The
constraint must include the \texttt{axis} (e.g., \texttt{x} or \texttt{y})
on which the elements should be ordered. The constraint must also specify
the node property \texttt{by} which the order is determined. The user can
optionally define an explicit list of values for a custom
\texttt{order}. The user may also optionally include a boolean of whether
or not to \texttt{reverse} the order. Users may finally specify an optional
\texttt{band} property for the constraint that determines a size for each
set region; without the \texttt{band} property, the first and last elements
in the order are free to move on either end, whereas the \texttt{band}
enforces them to fit within the same sized space as the intermediate
elements. \jheer{I don't totally follow what band is.}

\jheer{Similar question as above regarding x/y vs. other possible options.}

\emph{Rationale.} Unlike position constraints, order constraints allow the
user to specify relationships between different sets of nodes to specify
the overall node layout. These relationships are particular useful for
sorting and hierarchical layouts on the different sets of nodes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Circle Constraints}
\constraint{\texttt{circle} \texttt{around} \emph{``center''}} Circle
constraints allow the user to specify a ring layout for a set of
elements. The constraint must include which element the set elements should
be positioned around or may specify ``center'' for a generic center point
to be used.

\jheer{I'm confused regarding the center. What is a generic center point? Can one specify a x,y point or related guide? What does it mean to put an element in the center? Does this constraint work with sets of sets?}

\emph{Rationale.} Circular structures are a common layout requirement of
existing strategies \orange{citation} and can be crucial for relating
properties of the underlying structure.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Unit Constraints}
\constraint{\texttt{unit}} Unit constraints specify additional layout
properties for elements in the set. The user may optionally specify a
boolean \texttt{enclose} that prevents the set elements from overlapping
with other sets (\orange{figure}). The user may optionally specify an
\texttt{attract} property as either \texttt{center}, \texttt{left},
\texttt{right}, \texttt{top}, or \texttt{bottom} that causes the nodes to
be drawn to the specified area of the set (\orange{figure}). An optional
\texttt{padding} property determines the minimum amount of spacing that
should exist between elements in the unit and finally an optional
\texttt{spacing} property determines whether the spacing is either
\texttt{free} or \texttt{fixed}.

\todo{from Alan: ``attract'' to me connotes a soft constraint. But are
  these constraints actually required? If so, maybe call it ``alignment''
  or something else?} 
\todo{Jane: this section is pretty new, so probably want to discuss in the
meeting this week. The ``unit'' constraints here are supposed to help cover
some of the layout options that are a bit softer than the position and
alignment constraints otherwise discussed in this section. I am not sure
on the naming or anything here really. ``attract'' implies a kind of
organic clustering. Another option for this could be something like ``shape''
with options ``organic'' (the current attract) and ``grid.'' Maybe that would
be better, since I'm not sure it is overly necessary to have this sort of
alignment on the clustering.}

\jheer{We discussed this last week, and believe we decided to break this up into multiple constraints, right?}

\emph{Rationale.} The unit constraint allows the user to enforce more
control on how the nodes within a set behave outside of the general layout
constraints specified earlier. These constraints allow the user to apply
more nuanced behaviors between the nodes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Unsatisfiable Constraints in \projectname}

For a \projectname\ specification, it is possible to create node sets that
are \emph{not} disjoint and may thus be easily susceptible to the
specification of unsatisfiable constraints. Furthermore, through various
combinations of set specifications and constraint applications, it is
possible for the user to specify unsatisfiable constraints in the
\projectname\ layout. An example of a layout with unsatisfiable constraints
is shown in Figure \ref{fig:contradiction-example}. This layout completes
two partitions of the nodes (once by \texttt{level} and once by
\texttt{rank}); these collections lifted and an order constraint is applied
to enforce a hierarchy between the sets. However, due to the properties of
the nodes, the node with \texttt{rank 10} and \texttt{level 3} faces a
contradiction in the layout. Figure \ref{fig:contradiction-example}c shows
the error on the nodes that have constraints unsatisfied by the layout
(where darker means more error).

\todo{Does it make sense to talk about this here, or should we provide this
  as an example that we talk about later? Or should we talk about it more
  generally without an example and discuss why this is
  allowed?}

\jheer{This section notes that conflicts can arise, but not much else. What facilities if any does the language provide to help with this? How might a user learn that they've authored conflicting constraints? If this is more of a language runtime issue, we might note that there. For example at this point we haven't said anything about how our high-level constraints get translated into low-level constraints.}
