%!TEX root = human-constraint-layout.tex
\section{High-Level, User-Defined Constraints}
\treeLayout
In this section we describe our constraint language for high-level, user-defined constraints. This languages includes support for specifying sets of nodes over which to run the constraints and three types of constraints over the nodes: \emph{alignment}, \emph{order}, and \emph{position}. Figure \ref{fig:small-tree} and Figure \ref{fig:tree-layout} show an example specification for a tree layout and the resulting tree. 

We refer to two types of constraints throughout this paper: \emph{high-level} constraints and \emph{layout} constrains. \emph{High-level} constraints identify sets over which node-specific \emph{layout} constraints are applied. Every \emph{high-level} constraint must include a \texttt{"name"} property and specify sets over which to apply the \emph{layout} constraints. However, \emph{layout} constraints for a \emph{high-level} constraint are optional; in cases where \emph{layout} constraints are not provided, the user may simply want to define sets of nodes to refer to later in her constraint specification.

\subsection{Specifying Sets}
\label{sec:sets}
Every high-level constraint must specify a group of sets over which the layout constraints are defined. There are two types of high-level constraints that can be specified: \emph{within set} constraints and \emph{between set} constraints. \emph{Within set} constraints describe constraints that are applied to all of the nodes within a given set whereas \emph{between set} constraints describe constraints that are between larger sets. We describe how to define each type of constraint and show examples of how these constraints are applied.

\subsubsection{Within Set Constraints}
The user specifies within set constraints with the \texttt{"set"} property as shown for \texttt{alignLayer} in Figure \ref{fig:small-tree}. There are two ways to specify sets for within set constraints: the user can either partition all the nodes into disjoint sets or the user can specify expressions to create specific sets of nodes.

The simplest set definition is to \texttt{"partition"} nodes into sets based on a property of the node. When generating the sets, the system extracts the property of the node to use as the key for the set. Using \texttt{"partition"}, nodes can only occur in one set specification for the given constraint.

There are two additional properties that users can apply to set partitioning: \texttt{"include"} and \texttt{"ignore"}. \texttt{"include"} allows the user to specify a property of the accepted node to \emph{include} in the set. This property must also be a node, for example the \texttt{"parent"} or \texttt{"firstchild"} of the node (e.g. the \texttt{alignChildren} constraint of Figure \ref{fig:aligned-tree}). \texttt{"ignore"} represents a list of keys to \emph{ignore} when creating the sets. This property can be useful for preventing nodes with specific properties from being included in the layout (e.g. \texttt{orderNodes} in Figure \ref{fig:aligned-tree} and \texttt{layer} in Figure \ref{fig:tlr4-layout}).

Alternatively, the user can specify a concrete list of sets to compute. In this representation, the user defines an expression that returns a boolean value of whether or not the node should be included in the set. The user may also specify an optional \texttt{"name"} property that is used by the debugging tools. Using this syntax, users may specify sets such that a node can appear in multiple sets. However, we currently throw an error when such a specification occurs. Future work is required to determine whether or not this behavior would be useful for complex layouts. The \texttt{"expr"} definition currently supports value comparisons (e.g. \texttt{==}, \texttt{<=}, \texttt{>}), and conjunctions (\texttt{\&\&}) and disjunctions (\texttt{||}) of value comparisons and boolean values. The user may use \texttt{datum.property} to refer to properties of the node.

Within set constraints define constraints that should be applied to all nodes in a set separate from nodes outside the set. In Figure~\ref{fig:small-tree}, the user partitions nodes into sets based on their depth and creates an \emph{alignment} constraint to align the nodes in each set along the x-axis.

\subsubsection{Between Set Constraints}
The user specifies between set constraints using the \texttt{"from"} property as shown by the \texttt{orderLayers} constraint in Figure~\ref{fig:small-tree}. The user specifies between set constraints as a list of previously named constraints. Between set constraints define constraints that should be applied between sets but not between nodes in a given set. In Figure~\ref{fig:small-tree}, the user selects all the sets defined by \texttt{alignLayer} and creates an \emph{order} constraint to sort the sets by their depth along the y-axis.

\subsection{Specifying Constraints}
Once the user has defined sets of nodes over which to apply the constraints, she may optionally define layout constraints for that specification. Every cosntraint must have a \texttt{"type"} property that defines the type of constraint. We support three types of layout constraints: \emph{alignment} (\texttt{"align"}), \emph{order} (\texttt{"order"}), and \emph{position} (\texttt{"position"}).

\subsubsection{Alignment Constraints}
Alignment constraints are the easiest constraint to specify in our constraint language and define using WebCoLa. Alignment constraints must have two properties, \texttt{"type"} and \texttt{"axis"}. The property \texttt{"axis"} can be defined as either \texttt{"x"} or \texttt{"y"}. Figure~\ref{fig:small-tree} shows an example of \emph{alignment} for the high-level constraint \texttt{alignLayer}.

\subsubsection{Order Constraints}
Order constraints have three required properties: \texttt{"type"}, \texttt{"axis"} (either \texttt{"x"} or \texttt{"y"}), and \texttt{"by"}. The property \texttt{"by"} defines the property with which to order the nodes. If the \texttt{"by"} property does not have an obvious ordering (e.g. numeric, alphabetical), the user may optionally define an \texttt{"order"} property that is an ordered list of the expected inputs to the ordering function. The user may also optionally define a \texttt{"reverse"} property to reverse the behavior of the ordering. Figure~\ref{fig:small-tree} shows an example of an \emph{order} constraint for the high-level constraint \texttt{orderLayers}, which forces the layers to be positioned along the y-axis based on their depth.

\subsubsection{Position Constraints}
Position constraints have three required properties: \texttt{"type"}, \texttt{"position"}, and \texttt{"of"}. The property \texttt{"position"} accepts the values \texttt{"right"}, \texttt{"left"}, \texttt{"above"}, or \texttt{"below"}. The \texttt{"of"} property can be defined as a node, for example the \texttt{"parent"} or \texttt{"firstchild"} of the current node, or as a temporary point. The temporary point definition can include any combination of the properties \texttt{"name"}, \texttt{"x"}, and \texttt{"y"}. If \texttt{"x"} or \texttt{"y"} is not defined, it is initialized to zero by default. Specifying a \texttt{"name"} property allows the new node to be reused in different parts of the specification. The temporary nodes are optionally visualized in the layout and can be used to modify the final layout by moving the nodes. An example of this behavior is shown in Figure \ref{fig:serengeti-iterations}b,c.

\subsection{Built-in Properties}
The constraint specification can use any properties of the nodes in the original graph. However, we also provide a number of built-in properties that can be defined over the nodes. These properties are automatically computed and added to the graph specification prior to computing the final WebCoLa constraints. These properties are only added to nodes if such a property does not already exist on the nodes.

\begin{description}
\item[\texttt{\_id}] The index of the node in the graph specification.
\item[\texttt{depth}] One more than the max depth of the node's parents. This property is only allowed if the graph contains only one root node and does not contain cycles.
\item[\texttt{parent}] The parent of the current node. This property is only allowed if the node has one parent or is defined as the parent with the smallest \texttt{"\_id"}.
\item[\texttt{firstchild}] The child node of the current node with the smallest \texttt{"\_id"}.
\item[\texttt{parents}] The list of nodes that have edges where the current node is the target.
\item[\texttt{children}] The list of nodes that have edges where the current node is the source.
\item[\texttt{neighbors}] The list of nodes that have edges connected to the current node. This property is the join of the \texttt{"parents"} and \texttt{"children"} properties.
\item[\texttt{degree}] The number of \texttt{"neighbors"}.
\end{description}
